---
title: "Defining Custom Traits"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Defining Custom Traits}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette provides a comprehensive guide for defining custom derived traits using `glydet`.
Before proceeding, please ensure you have read the "Get Started with glydet" vignette
and are familiar with the fundamental concepts of derived traits and meta-properties.

**Note about new features:**

We introduced an experimental function `make_trait()` since glydet 0.7.0.
This function allows you to create a trait from natural language.
Check out the last section of this vignette for details.

```{r setup}
#| message: false

library(dplyr)
library(glydet)
library(glyexp)
library(glyclean)
library(glyrepr)

exp <- auto_clean(real_experiment)
```

## Custom Traits

`glydet` provides three trait factory functions for creating custom derived traits:

- `prop()` for calculating the abundance proportion of a specific glycan subset
- `ratio()` for computing the abundance ratio between two glycan subsets
- `wmean()` for calculating the weighted mean of a quantitative property, weighted by glycan abundances

All derived traits in `glydet` are constructed using these three factory functions.
The definitions of all built-in traits demonstrate their usage:

- `TM`: `prop(Tp == "highmannose")`
- `TH`: `prop(Tp == "hybrid")`
- `TC`: `prop(Tp == "complex")`
- `MM`: `wmean(nM, within = (Tp == "highmannose"))`
- `CA2`: `prop(nA == 2, within = (Tp == "complex"))`
- `CA3`: `prop(nA == 3, within = (Tp == "complex"))`
- `CA4`: `prop(nA == 4, within = (Tp == "complex"))`
- `TF`: `prop(nF > 0)`
- `TFc`: `prop(nFc > 0)`
- `TFa`: `prop(nFa > 0)`
- `TB`: `prop(B)`
- `GS`: `wmean(nS / nG)`
- `AG`: `wmean(nG / nA)`
- `TS`: `prop(nS > 0)`

These definitions utilize meta-properties as building blocks.
For instance, `T` represents glycan type, `nM` denotes the number of mannose residues, and so forth.
To retrieve all available built-in meta-properties, use `names(all_mp_fns())`.

The complete list of built-in meta-properties is provided below:

| Name | Function | Type | Description |
|------|----------|------|-------------|
| `T` | `n_glycan_type()` | factor | Type of the glycan, either "complex", "hybrid", "highmannose", or "pausimannose" |
| `B` | `has_bisecting()` | logical | Whether the glycan has a bisecting GlcNAc |
| `nA` | `n_antennae()` | integer | Number of antennae |
| `nF` | `n_fuc()` | integer | Number of fucoses |
| `nFc` | `n_core_fuc()` | integer | Number of core fucoses |
| `nFa` | `n_arm_fuc()` |integer | Number of arm fucoses |
| `nG` | `n_gal()` | integer | Number of galactoses |
| `nGt` | `n_terminal_gal()` | integer | Number of terminal galactoses |
| `nS` | `n_sia()` | integer | Number of sialic acids |
| `nM` | `n_man()` | integer | Number of mannoses |

The following sections provide detailed explanations of each factory function.

### Trait Factories

#### `prop()`

`prop()` creates proportion traits, which represent the most common type of derived traits.
These traits calculate the relative abundance of a specific glycan subset within a defined population.
Examples include the proportion of core-fucosylated glycans within all glycans,
or the proportion of tetra-antennary glycans within the complex glycan subset.

`prop()` accepts an expression that evaluates to a logical vector.
Both built-in and custom meta-properties (covered later) can be referenced within the expression.

For example, the proportion of core-fucosylated glycans within all glycans is defined as:

```r
prop(nFc > 0)
```

Since `nFc` is an integer meta-property representing the number of core fucoses,
the expression `nFc > 0` evaluates to a logical vector, creating a valid trait definition.

Consider this simpler example:

```r
prop(B)
```

This demonstrates a straightforward trait definition.
Since `B` (bisecting GlcNAc presence) is already a logical meta-property,
`prop(B)` constitutes a valid trait definition.

A more complex example demonstrates compound logical operations:

```r
prop(nS > 0 & nFa > 0)
```

This trait calculates the proportion of glycans containing both sialic acid and arm fucose.
Both `nS > 0` and `nFa > 0` represent logical expressions,
combined using the logical AND operator (`&`).
Any R logical operator (including `|`, `!`, etc.) can be utilized within these expressions.

Arithmetic calculations can also be incorporated within expressions.
For example, the `CF` trait is defined as:

```r
prop(nF > 0)
```

We know that `nFc + nFa` is equivalent to `nF`, so the above definition is equivalent to:

```r
prop((nFc + nFa) > 0)
```

As a general principle, any R expression that evaluates to a logical vector is valid for use with `prop()`.

The traits described above calculate proportions relative to the entire glycan population.
To customize the denominator (reference population), the `within` parameter can be employed.
For example, calculating the proportion of bi-antennary glycans within the **complex** glycan subset (`CA2`) requires this approach.

The `within` parameter accepts an expression that evaluates to a logical vector,
following the same syntax as the primary parameter.

The `CA2` trait is defined as follows:

```r
prop(nA == 2, within = (Tp == "complex"))
```

Note that parentheses around `Tp == "complex"` are optional.
Consequently, this alternative definition is equally valid:

```r
prop(nA == 2, within = Tp == "complex")
```

Combining these two parameters enables the creation of sophisticated trait definitions.
For instance, the proportion of sialylated glycans within the subset of core-fucosylated tetra-antennary glycans is defined as:

```r
prop(nS > 0, within = (nFc > 0 & nA == 4))
```

Having mastered the usage of `prop()`,
examine the definitions of built-in proportion traits to reinforce your understanding.

Practice exercises:

1. Define a trait calculating the proportion of glycans containing terminal galactose.
2. Define a trait calculating the proportion of glycans with terminal galactose but no sialic acid.
3. Define a trait calculating the proportion of glycans with exactly two antennae and no bisecting GlcNAc.
4. Define a trait calculating the proportion of glycans with bisecting GlcNAc within the bi-antennary glycan subset.

Solutions are provided at the end of this vignette.

#### `ratio()`

`ratio()` creates ratio traits that represent the quotient of total abundances between two glycan groups.
Examples include the ratio of complex to hybrid glycans,
or the ratio of bisecting to non-bisecting glycans.

`ratio()` accepts two expressions that evaluate to logical vectors, similar to `prop()`.

For example, the ratio of complex to hybrid glycans is defined as:

```r
ratio(Tp == "complex", Tp == "hybrid")
```

The first expression defines the numerator, while the second expression defines the denominator.

The `within` parameter can be employed to apply restrictions to both numerator and denominator.
For example, the ratio of bisecting to non-bisecting glycans within the bi-antennary subset is defined as:

```r
ratio(B, !B, within = (nA == 2))
```

This represents syntactic sugar for the more verbose:

```r
ratio(B & (nA == 2), (!B) & (nA == 2))
```

The `within` parameter provides clearer semantics and reduces redundancy.

Note that `prop()` represents a special case of `ratio()`,
where `prop(cond, within)` is mathematically equivalent to `ratio(cond & within, within)`.
The following two definitions are functionally identical:

```r
# using prop()
prop(nFc > 0, within = (Tp == "complex"))

# using ratio()
ratio(nFc > 0 & Tp == "complex", Tp == "complex")
```

However, `prop()` is recommended when appropriate for enhanced readability,
as it provides more intuitive semantics that align with natural language.
The `prop()` example above can be interpreted as "the proportion of core-fucosylated glycans within all complex glycans,"
whereas the `ratio()` version requires additional cognitive processing.

Practice exercises:

1. Define a trait calculating the ratio of core-fucosylated to non-core-fucosylated glycans.
2. Define a trait calculating the ratio of tetra-antennary to tri-antennary glycans within the complex glycan subset.
3. Redefine the `CA2` trait using `ratio()`.

#### `wmean()`

`wmean()` creates weighted-mean traits, which represent the most sophisticated yet powerful type of derived traits.
These traits calculate the abundance-weighted average of quantitative properties across glycan populations.
Examples include the average number of antennae across all glycans,
or the average degree of sialylation per galactose across the entire glycan repertoire.

Before exploring `wmean()` usage, it is essential to understand the weighted-mean concept.

Consider three glycans: `G1`, `G2`, and `G3`.
Their respective antennae counts are 1, 2, and 3,
with relative abundances of 50%, 20%, and 30%.
The weighted-mean number of antennae is calculated as:

```{r}
(1 * 0.5 + 2 * 0.2 + 3 * 0.3) / (0.5 + 0.2 + 0.3)
```

This value represents the abundance-weighted average degree of branching across the glycan population.

This concept provides substantial analytical power,
enabling the calculation of numerous biologically meaningful traits.

Consider another example: the average degree of sialylation per antenna across all glycans.
Using the same three glycans (`G1`, `G2`, and `G3`) with antennae counts of 1, 2, and 3,
sialic acid counts of 1, 1, and 3, respectively,
and relative abundances of 50%, 20%, and 30%.

First, calculate the sialylation degree per antenna for each glycan
(sialic acid count divided by antenna count):

```{r}
c(1/1, 1/2, 3/3)
```

Subsequently, incorporate the abundance weighting:

```{r}
(1/1 * 0.5 + 1/2 * 0.2 + 3/3 * 0.3) / (0.5 + 0.2 + 0.3)
```

This yields the abundance-weighted average degree of sialylation per antenna across the glycan population.

Having established the weighted-mean concept
and mastered `prop()` and `ratio()` fundamentals,
we can now examine `wmean()` implementation.

`wmean()` accepts expressions that evaluate to numeric vectors,
contrasting with the logical vectors required by `prop()` and `ratio()`.

For example, the average number of antennae across all glycans is defined as:

```r
wmean(nA)
```

The average degree of sialylation per antenna across all glycans is defined as:

```r
wmean(nS / nA)
```

The `within` parameter can be utilized to restrict the weighted-mean calculation to specific glycan subsets.
For example, the average number of antennae within complex glycans is defined as:

```r
wmean(nA, within = (Tp == "complex"))
```

Practice exercises:

1. Define a trait calculating the average degree of sialylation per antenna.
2. Define a trait calculating the average number of arm fucoses.

#### `total()`

`total()` creates total abundance traits, which is the sum of the abundances of a group of glycans.
This is the simplest type of derived traits in `glydet`.

For example, the total abundance of all complex glycans is defined as:

```r
total(Tp == "complex")
```

There is no `within` parameter for `total()`,
as you can always add a restriction to the expression by using `&`.

```r
total(Tp == "complex" & nA == 4)
```

#### `wsum()`

`wsum()` is the cousin of `wmean()`,
but instead of calculating the average value, it calculates the sum of the values.

Let's see an example to understand the difference between `wmean()` and `wsum()`.
Say we have three glycans: `G1`, `G2`, and `G3` in two samples: `S1` and `S2`.

- Sample S1: `G1` (10), `G2` (20), `G3` (30) (intensity values)
- Sample S2: `G1` (20), `G2` (40), `G3` (60) (intensity values)

```{r}
expr_mat <- matrix(c(10, 20, 30, 20, 40, 60), nrow = 3)
rownames(expr_mat) <- c("G1", "G2", "G3")
colnames(expr_mat) <- c("S1", "S2")
expr_mat
```

For simplicity, we don't use any meta-properties here, just the number 1.

```{r}
trait1 <- wmean(1)
trait1(expr_mat, tibble::tibble())
```

```{r}
trait2 <- wsum(1)
trait2(expr_mat, tibble::tibble())
```

See the difference?
`wmean()` calculates the average value of some numeric property,
so as long as the ratio of each glycan's abundance is the same,
the result is always the same.
But `wsum()` is sensitive to the abundance of each glycan,
so the result is different.

### Using Custom Traits

Having learned how to define custom traits, we can now implement them.

`derive_traits()` includes a `trait_fns` parameter that accepts a named list of derived trait functions.
Note that the three factory functions described above return functions as their output.

```{r}
class(prop(nFc > 0))
```

R supports functional programming paradigms, treating functions as first-class objects.
Functions can be passed as arguments to other functions and returned from function calls.

While understanding this concept enhances R proficiency,
it is not prerequisite for usage—simply define traits as a named list and provide it to the `trait_fns` parameter.

The following example examines sialylation degree within glycan subsets of varying antenna counts:

```r
my_traits <- list(
  A2S = wmean(nS / nA, within = (nA == 2)),
  A3S = wmean(nS / nA, within = (nA == 3)),
  A4S = wmean(nS / nA, within = (nA == 4))
)
derive_traits(exp, trait_fns = my_traits)
```

The identifiers "A2S", "A3S", and "A4S" represent the derived trait names.

When `trait_fns` is omitted, `derive_traits()` internally invokes `basic_traits()` to utilize built-in trait functions.

### Validating Trait Definitions

How can you ensure your trait definitions are accurate and meaningful?
The `explain_trait()` function provides an intuitive way to validate your trait definitions by generating human-readable explanations.

This function helps you:

- Verify that your trait logic matches your intended analysis
- Understand complex trait expressions at a glance  
- Debug and refine trait definitions before deployment

Let's examine the `A2S` trait defined earlier as an example:

```{r}
explain_trait(wmean(nS / nA, within = (nA == 2)))
```

The function interprets your trait expression and returns a clear,
natural language description of what the trait calculates.
This validation step is particularly valuable when working with complex trait definitions
or when collaborating with team members who need to understand your analytical approach.

We have introduced a new argument `use_ai` to `explain_trait()` since glydet 0.7.0.
If `use_ai` is TRUE, the function will use a Large Language Model (LLM) to explain the trait,
which covers more complex cases and edge cases.
Check out the documentation of `explain_trait()` for details.

## Custom Meta-Properties

The preceding examples utilized exclusively built-in meta-properties.
As the building blocks of derived traits,
understanding how to define custom meta-properties will bring you more flexibility.

`glydet` provides two means of defining custom meta-properties:

- By providing meta-property functions to the `mp_fns` parameter of `derive_traits()`.
- By using columns in the variable information tibble as meta-properties.

### Defining Custom Meta-Properties with Functions

Meta-property functions must accept a `glyrepr::glycan_structure()` vector
and return a vector of corresponding meta-property values.

For example, the built-in meta-property `B` is defined as follows:

```r
# Simplified version for illustration purposes
function(glycans) {
  motif <- "HexNAc(??-?)Hex(??-?)HexNAc(??-?)HexNAc(??-"
  glymotif::have_motif(glycans, motif, alignment = "core")
}
```

This implementation utilizes `glymotif::have_motif()` to determine motif presence within glycan structures.
Meta-property function definitions offer considerable flexibility.
Any tools from the `glycoverse` ecosystem or broader R environment can be employed,
provided the function accepts a `glyrepr::glycan_structure()` vector and returns an atomic vector.
All built-in meta-properties utilize the `glymotif` package for motif matching and counting operations.
For detailed information, consult the [glymotif documentation](https://glycoverse.github.io/glymotif/).

The following example demonstrates custom meta-property creation:

```{r}
my_mp_fns <- list(
  # Number of Lewis antigens
  nLe = ~ glymotif::count_motif(.x, "Hex(??-?)[dHex(??-?)]HexNAc(??-"),
  # Number of poly-LacNAc units
  nPl = ~ glymotif::count_motif(.x, "Hex(??-?)HexNAc(??-?)Hex(??-?)HexNAc(??-")
)
```

Note that purrr-style lambda functions provide a concise syntax for meta-property function definitions.

Traits can be defined using these new meta-properties:

```{r}
my_traits <- list(
  # Average number of Lewis antigens per antenna
  LeA = wmean(nLe / nA),
  # Average number of poly-LacNAc units per glycan
  Pl = wmean(nPl)
)
```

These expressions incorporate both custom meta-properties (`nLe` and `nPl`) and the built-in meta-property `nA`.

Implementation of these custom meta-properties and traits proceeds as follows:

```{r}
derive_traits(exp, trait_fns = my_traits, mp_fns = c(my_mp_fns, all_mp_fns()))
```

Ensure that custom meta-properties are combined with built-in meta-properties
when the latter are required within trait expressions.

Let's see another example about sialic acid linkage types.
Neu5Ac can have two different linkage types: a2-3 and a2-6.
We can define two custom meta-properties to count the number of sialic acids with each linkage type:

```{r}
sia_mp_fns <- list(
  # Number of a2-3 sialic acids
  nL = ~ glymotif::count_motif(.x, "NeuAc(a2-3)Hex(??-"),
  # Number of a2-6 sialic acids
  nE = ~ glymotif::count_motif(.x, "NeuAc(a2-6)Hex(??-")
)
```

And define two traits to calculate the degree of a2-3 and a2-6 sialylation per galactose:

```{r}
sia_traits <- list(
  # Average degree of a2-3 sialylation per galactose
  GL = wmean(nL / nG),
  # Average degree of a2-6 sialylation per galactose
  GE = wmean(nE / nG)
)
```

And calculate the traits:

```{r}
derive_traits(exp, trait_fns = sia_traits, mp_fns = c(sia_mp_fns, all_mp_fns()))
```

This is an example of how you can violate the [ambiguity assumption](https://glycoverse.github.io/glydet/articles/glydet.html#working-with-structural-ambiguity)
of built-in meta-properties and derived traits,
by introducing more sophisticated and linkage-awared meta-properties.

### Defining Custom Meta-Properties with Columns

The first method is handy if all information you need is already encoded in glycan structures.
However, there are situations that you need to rely on other meta-data.
In this case, you can use columns in the variable information tibble as meta-properties directly,
by specifying the `mp_cols` parameter.

For example, let's use the sialic acid linkage type example above.
You might have used special derivatization methods to differentiate the two linkage types.
This information might end up in a column in the variable information tibble,
not directly in the glycan structures.

```{r}
# Here we assume all sialic acids are a2-6
exp2 <- exp |>
  mutate_var(
    n_a26_sia = count_mono(glycan_structure, "NeuAc"),
    n_a23_sia = 0L
  )
```

```{r}
exp2 |>
  get_var_info() |>
  filter(n_a26_sia > 0) |>
  pull(glycan_structure)
```

See? No linkage information about sialic acids is in the glycan structures.
But we have two columns: `n_a26_sia` and `n_a23_sia` storing this information.

Now we can define two traits to calculate the degree of a2-3 and a2-6 sialylation per galactose:

```{r}
sia_traits <- list(
  # Average degree of a2-3 sialylation per galactose
  GL = wmean(nL / nG),
  # Average degree of a2-6 sialylation per galactose
  GE = wmean(nE / nG)
)
```

And use `mp_cols` to tell `derive_traits()` to use these columns as meta-properties:

```{r}
derive_traits(exp2, trait_fns = sia_traits, mp_cols = c(nL = "n_a23_sia", nE = "n_a26_sia"))
```

## Using `make_trait()`

`make_trait()` allows you to create a trait from natural language.
To use this feature, you need to install the `ellmer` package (not a forced dependency of `glydet`).
You also need to provide an API key for the DeepSeek chat model.
You can obtain an API key from https://platform.deepseek.com.
Please set the environment variable `DEEPSEEK_API_KEY` to your API key using `Sys.setenv()`:

```r
# Works for a single session
Sys.setenv(DEEPSEEK_API_KEY = "your_api_key")
```

Then you can transform your ideas into trait functions by calling `make_trait()`:

```r
my_traits <- list(
  nS = make_trait("the average number of sialic acids"),
  nG = make_trait("the average number of galactoses")
)
derive_traits(exp, trait_fns = my_traits)
```

This function is experimental, so the result should be verified manually.
If the description is not clear, an error will be raised.
Try to read the descriptions of built-in traits to get ideas.

## Exercise Solutions

**`prop()`**

1. `prop(nGt > 0)`
2. `prop(nGt > 0 & nS == 0)`
3. `prop(nA == 2 & !B)`
4. `prop(B, within = (nA == 2))`

**`ratio()`**

1. `ratio(nFc > 0, nFc == 0)`
2. `ratio(nA == 4, nA == 3, within = (Tp == "complex"))`
3. `ratio(nA == 2 & Tp == "complex", Tp == "complex")`

**`wmean()`**

1. `wmean(nS / nA)`
2. `wmean(nFa)`
